---
title: "Urban Expansion over Disaster-Susceptible Areas in Brazil (1985-2025)"
author: "Analysis Script"
format: html
execute:
  warning: false
  message: false
---

## 0. Setup and Libraries

First, we load the necessary packages. We will use `tidyverse` for data manipulation, `sf` for vector spatial data, `terra` for heavy raster processing, and `geobr` to fetch Brazilian municipalities.

```{r}
'# Install pak first
install.packages("pak")

# Use pak to install your packages in a fraction of the time
pak::pkg_install(c("tidyverse", "sf", "terra", "fs", "geobr"))'
```

```{r setup}
# Load required libraries
library(tidyverse)
library(sf)
library(terra)
library(geobr)
library(fs)


# Ensure temp directory exists
if (!dir_exists("data/temp")) dir_create("data/temp")

# Optimize terra memory usage (use only 50% of available RAM before swapping to disk)
terraOptions(memfrac = 0.5, tempdir = "data/temp")

```

## 1. Create Data Directory and Gitignore
We create the data directory to store heavy spatial files and ensure it is ignored by Git to prevent committing large files.

```{r}
# Define data directory path
data_dir <- "data"

# Create directory if it doesn't exist
if (!dir_exists(data_dir)) {
  dir_create(data_dir)
}

# Add data folder to .gitignore
gitignore_path <- ".gitignore"
ignore_entry <- "data/"

if (!file_exists(gitignore_path) || !any(str_detect(read_lines(gitignore_path), ignore_entry))) {
  write_lines(ignore_entry, gitignore_path, append = TRUE)
}
```



## 2. Download MapBiomas Raster
We check if the MapBiomas GeoTIFF (Collection 11) is already present. If not, we download it directly from Google Cloud Storage.


```{r}
# Define the target years for our analysis
target_years <- c(1985, 2025)

# Create a character vector of local file paths
mapbiomas_files <- path(data_dir, str_glue("brazil_coverage-col11_{target_years}.tif"))

# Iterate over the years and download if missing
for (i in seq_along(target_years)) {
  yr <- target_years[i]
  file_path <- mapbiomas_files[i]
  
  if (!file_exists(file_path)) {
    url <- str_glue("[https://storage.googleapis.com/mapbiomas-public/initiatives/brasil/collection11/lulc/coverage/brazil_coverage/brazil_coverage-col11](https://storage.googleapis.com/mapbiomas-public/initiatives/brasil/collection11/lulc/coverage/brazil_coverage/brazil_coverage-col11)_{yr}.tif")
    
    download.file(
      url = url,
      destfile = file_path,
      mode = "wb"
    )
  }
}
```

## 3. Download Susceptibility Shapefiles (Geological Survey of Brazil)
We establish a helper function to read Web Feature Service (WFS) data from the SGB URLs, extracting the base URL from the GetCapabilities string, reading the spatial layer, and saving it locally as a Shapefile to avoid repeated heavy requests.


```{r}
'# SGB WFS URLs
url_inundation <- "[https://geoportal.sgb.gov.br/server/services/gestaoterritorial/inundacao/MapServer/WFSServer](https://geoportal.sgb.gov.br/server/services/gestaoterritorial/inundacao/MapServer/WFSServer)"
url_flash_flood <- "[https://geoportal.sgb.gov.br/server/services/gestaoterritorial/enxurrada/MapServer/WFSServer](https://geoportal.sgb.gov.br/server/services/gestaoterritorial/enxurrada/MapServer/WFSServer)"
url_debris_flow <- "[https://geoportal.sgb.gov.br/server/services/gestaoterritorial/corrida_de_massa/MapServer/WFSServer](https://geoportal.sgb.gov.br/server/services/gestaoterritorial/corrida_de_massa/MapServer/WFSServer)"


# Helper function to download and read WFS layers
fetch_wfs_data <- function(base_url, file_name) {
  file_path <- path(data_dir, file_name)
  
  if (!file_exists(file_path)) {
    # Prefix with WFS: for GDAL/sf
    wfs_connection <- str_c("WFS:", base_url)
    
    # Automatically get the first layer name available in the service
    layer_info <- st_layers(wfs_connection)
    target_layer <- layer_info$name[1]
    
    # Read from WFS and save locally
    wfs_data <- st_read(wfs_connection, layer = target_layer)
    st_write(wfs_data, file_path, driver = "ESRI Shapefile")
  }
  
  # Read the local shapefile
  st_read(file_path, quiet = TRUE)
}

# Fetch all four datasets
inundation <- fetch_wfs_data(url_inundation, "inundation.shp")
flash_flood <- fetch_wfs_data(url_flash_flood, "flash_flood.shp")
debris_flow <- fetch_wfs_data(url_debris_flow, "debris_flow.shp")
mass_movement <- fetch_wfs_data(url_mass_movement, "mass_movement.shp")'
```

## 4. Create Susceptibility Overlay Shapefile
We bind all four spatial data frames into a single object and create an overlaid spatial union. The result is saved as a new Shapefile.
```{r create-overlay}
# Define local paths (removed leading slash to ensure relative path works cross-platform)
debris_path <- path(data_dir, "sig_angradosreis_rj/Suscetibilidade/Corrida_de_Massa_A.shp")
inundation_path <- path(data_dir, "sig_angradosreis_rj/Suscetibilidade/Inundacao_A.shp")
flash_flood_path <- path(data_dir, "sig_angradosreis_rj/Suscetibilidade/Enxurrada_A.shp")

# Read local shapefiles
debris_flow <- st_read(debris_path, quiet = TRUE)
inundation <- st_read(inundation_path, quiet = TRUE)
flash_flood <- st_read(flash_flood_path, quiet = TRUE)

# Combine polygons into a single susceptibility area for the masking process
target_crs <- st_crs(inundation)

susceptibility_area <- bind_rows(
  inundation |> st_transform(target_crs) |> select(geometry),
  flash_flood |> st_transform(target_crs) |> select(geometry),
  debris_flow |> st_transform(target_crs) |> select(geometry)
) |> 
  st_make_valid()


```

## 5. List Municipalities Overlapped by Susceptibility Area
We use the geobr package to retrieve all Brazilian municipalities and intersect them with the unified susceptibility shapefile. The resulting list is exported to a data frame.


```{r}
'# Download Brazilian municipalities (using 2022 census data as baseline)
municipalities <- read_municipality(year = 2022, showProgress = FALSE)

# Transform to match the susceptibility CRS
municipalities <- st_transform(municipalities, st_crs(susceptibility_area))

# Keep only municipalities that intersect with the susceptibility area
overlapped_munis <- st_filter(municipalities, susceptibility_area)

# Convert to standard data.frame (dropping geometry)
overlapped_df <- overlapped_munis |> 
  st_drop_geometry() |> 
  as_tibble()

# Preview the data
head(overlapped_df)'
```

## 5. Sequential Raster Processing and Masking
We will iterate through each year sequentially. We first crop the raster to the extent of the susceptibility areas (very fast operation) and check where it became urbanized (Class 24). If it's the first time a pixel became urban, we record that year in our master raster first_urban. Finally, we mask this combined raster perfectly to the polygons.

```{r process-raster-efficiently}
urban_expansion_path <- path(data_dir, "urban_expansion.tif")

if (!file_exists(urban_expansion_path)) {
  
  # Read the first raster merely to extract its CRS for projecting our vector
  template_raster <- rast(mapbiomas_files[1])
  
  susceptibility_vect <- vect(susceptibility_area) |> 
    project(crs(template_raster))
  
  rm(template_raster)
  gc()
  
  # Initialize an empty object to store the temporal urban compilation
  first_urban <- NULL
  
  # Iterate sequentially through the years
  for (i in seq_along(target_years)) {
    yr <- target_years[i]
    file_path <- mapbiomas_files[i]
    
    # 1. Load the raster and crop it strictly to the vector's bounding box
    mb_raster <- rast(file_path)
    mb_cropped <- crop(mb_raster, susceptibility_vect)
    
    # 2. Create boolean raster representing class 24 (Urban Area)
    is_urban <- mb_cropped == 24
    
    # 3. Consolidate into the master map
    if (is.null(first_urban)) {
      # For the first year (1985), if urban, assign year, else NA
      first_urban <- ifel(is_urban, yr, NA)
    } else {
      # For subsequent years, update pixel to current year ONLY IF:
      # - The pixel is currently urban (is_urban == TRUE)
      # - AND the pixel was NOT previously marked as urban (is.na(first_urban))
      first_urban <- ifel(is_urban & is.na(first_urban), yr, first_urban)
    }
    
    # Clean up intermediate heavy objects per iteration
    rm(mb_raster, mb_cropped, is_urban)
    gc()
  }
  
  # 4. Mask ONLY the final consolidated 1-band raster
  # This sets pixels completely outside the susceptibility polygons to NA
  final_urban_expansion <- mask(first_urban, susceptibility_vect)
  
  # 5. Save the output
  writeRaster(
    final_urban_expansion, 
    urban_expansion_path, 
    overwrite = TRUE, 
    datatype = "INT2U"
  )
}

# Clear vector from RAM
rm(susceptibility_area)
gc()

# Load and plot final result
urban_expansion <- rast(urban_expansion_path)

```

## 6. Estimate Expansion Area by Category and Municipality

To calculate the exact area of urban expansion over each susceptibility category per municipality, we avoid vector intersections entirely. Instead, we rasterize the municipality polygons using a unique ID, stack it with our urban expansion map, and perform a cross-tabulation. This is orders of magnitude faster than spatial polygon extractions.

```{r calculate-area-muni}
# Load the master urban expansion raster
urban_expansion_path <- path(data_dir, "urban_expansion.tif")
urban_expansion <- rast(urban_expansion_path)

# 1. Prepare municipality vector data and create a numeric ID lookup table
# 'overlapped_df' and 'overlapped_munis' come from Step 4
muni_info <- overlapped_df |> 
  mutate(muni_id = row_number())

muni_vect <- overlapped_munis |> 
  mutate(muni_id = row_number()) |> 
  vect() |> 
  project(crs(urban_expansion))

# 2. Rasterize municipalities using the unique numeric ID
# This creates a raster where each pixel value is the municipality ID
muni_raster <- rasterize(
  muni_vect, 
  urban_expansion, 
  field = "muni_id", 
  background = NA
)

# Define the categories and their respective shapefile paths
susceptibility_files <- list(
  "Inundation" = "inundation.shp",
  "Flash Flood" = "flash_flood.shp",
  "Debris Flow" = "debris_flow.shp",
  "Mass Movement" = "mass_movement.shp"
)

# Initialize an empty list to store the results
area_results_muni <- list()

# 3. Loop through each category to perform Crosstabulation
for (cat_name in names(susceptibility_files)) {
  
  # Load and project the vector data for the specific category
  shp_path <- path(data_dir, susceptibility_files[[cat_name]])
  cat_vect <- vect(shp_path) |> 
    project(crs(urban_expansion))
  
  # Rasterize the category polygons
  cat_raster <- rasterize(
    cat_vect, 
    urban_expansion, 
    field = 1, 
    background = NA, 
    touches = TRUE
  )
  
  # Mask the urban expansion map keeping ONLY pixels inside this category
  urban_in_cat <- mask(urban_expansion, cat_raster)
  
  # Rename layers to ensure clean crosstab output column names
  names(muni_raster) <- "muni_id"
  names(urban_in_cat) <- "year"
  
  # Create a multi-band raster stack (Layer 1: Muni ID, Layer 2: Year)
  stacked_rasters <- c(muni_raster, urban_in_cat)
  
  # Crosstabulate to count frequencies of unique combinations (muni_id vs year)
  # 'long = TRUE' returns a data.frame
  # 'useNA = FALSE' ignores pixels that are empty in either layer (crucial for RAM)
  cross_tab <- crosstab(stacked_rasters, long = TRUE, useNA = FALSE) |> 
    as_tibble() |> 
    rename(pixels = Freq)
  
  # Calculate area metrics and join municipality names back via muni_id
  muni_cat_summary <- cross_tab |> 
    mutate(
      category = cat_name,
      area_m2 = pixels * 900,
      area_ha = area_m2 / 10000,
      area_km2 = area_m2 / 1e6
    ) |> 
    left_join(
      muni_info |> select(muni_id, code_muni, name_muni, name_state, abbrev_state), 
      by = "muni_id"
    ) |> 
    select(
      code_muni, name_muni, abbrev_state, category, year, pixels, 
      area_m2, area_ha, area_km2
    )
  
  # Store in the list
  area_results_muni[[cat_name]] <- muni_cat_summary
  
  # Free up RAM before the next loop iteration
  rm(cat_vect, cat_raster, urban_in_cat, stacked_rasters, cross_tab)
  gc()
}

# 4. Bind all lists into a single tidy dataframe
final_expansion_muni <- bind_rows(area_results_muni) |> 
  arrange(abbrev_state, name_muni, category, year)

# Clear massive spatial objects from memory
rm(muni_raster, muni_vect, urban_expansion)
gc()

# Preview the final data
head(final_expansion_muni)

```